Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-14 manuscript_sciad.Rmd
edits: 'Supp Text' to "Supp Mat"

Given the results of extensive sensitivity analysis (**Supplementary Text**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario reproduces natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary[@petersonModellingDendriticEcological2013; @mcguireNetworkAnalysisReveals2014]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities [@poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability $p_{d}$, and the maximum value of interspecific competition strength $\alpha_{max}$ (see **Materials and Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Text**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges (**Figure 1D**) but support distinct fish communities (**Tables S5 and S6**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Materials and Methods** and **Supplementary Text**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Materials and Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at a multi-generation time scale [@comteFishDispersalFlowing2018], while posing challenges to traverse across watersheds (the ocean or lentic habitats; see **Materials and Methods** for watershed definition). We combined this data set with geospatial information, including watershed characteristics (watershed area and branching probability), climates (annual mean temperature and annual cumulative precipitation), and land use patterns (the fraction of agricultural land use and dam density). Using this data set, we assessed whether observed effects of the macro-scale factors are consistent across the two geographic regions by developing global and region-specific robust regression models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Materials and Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Materials and Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions. The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Text**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual control of $\gamma$ diversity to emerge. Further, the consistent patterns across geographic regions with distinct fish fauna suggest that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

We used `r n_scenario` sets of parameter combinations (i.e., simulation scenarios) based on the results of extensive sensitivity analysis (**Supplementary Text**). The parameter combinations comprises four landscape and eight ecological scenarios. Landscape scenarios distinguish spatial patterns of habitat heterogeneity, while ecological scenarios change ecological traits of simulated species. For landscape scenarios, we used four combinations of environmental variation at headwaters ($\sigma_h = 0.01, 1$) and the degree of local environmental noise ($\sigma_l = 0.01, 1$). The inequality of $\sigma_h$ and $\sigma_l$ controls the relationship between branching complexity and habitat heterogeneity (**Figures 1A and 1B**). For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

@@

We compiled fish data at two geographic regions: the Hokkaido island, Japan, and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish [@fukushimaModellingEffectsDams2007; @fukushimaHokkaidoFreshwaterFish2011], monitoring data at protected watersheds [@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors [@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire island. For the Midwest, we assembled fish community data collected by the Iowa Department of Natural Resources [@iowadepartmentofnaturalresourcesBiologicalAssessmentIowa2004], Illinois Environmental Protection Agency and Illinois Department of Natural Resources [@illinoisenvironmentalprotectionagencyIllinoisWaterMonitoring2014], Minnesota Pollution Control Agency [@minnesotapollutioncontrolagencyDevelopmentFishbasedIndex2014], and Wisconsin Department of Natural Resources [@wisconsindepartmentofnaturalresourcesGuidelinesAssessingFish2018]. We detailed procedures for initial data selection in **Supplementary Text**.

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating asymptotic species richness (**Supplementary Text**).

@@

For the supported model of $\gamma$ diversity, we estimated average predictive comparisons of watershed area and branching probability to compare their effect sizes. The average predictive comparison is an expected change of the response variable per a unit difference of the explanatory variable of interest [@gelmanAveragePredictiveComparisons2007]. Unlike other common metrics of effect size (e.g., standardized regression coefficients), this method provides intuitive yet rigorous estimates of effect size in models with non-linear variable transformations (see **Supplementary Text** for further details). We performed statistical analysis using R version 4.0.2 [@rcoreteamLanguageEnvironmentStatistical2020].

>>>>

Given the results of extensive sensitivity analysis (**Supplementary Materials**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario reproduces natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary[@petersonModellingDendriticEcological2013; @mcguireNetworkAnalysisReveals2014]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities [@poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability $p_{d}$, and the maximum value of interspecific competition strength $\alpha_{max}$ (see **Materials and Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Materials**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges (**Figure 1D**) but support distinct fish communities (**Tables S5 and S6**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Materials and Methods** and **Supplementary Materials**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Materials and Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at a multi-generation time scale [@comteFishDispersalFlowing2018], while posing challenges to traverse across watersheds (the ocean or lentic habitats; see **Materials and Methods** for watershed definition). We combined this data set with geospatial information, including watershed characteristics (watershed area and branching probability), climates (annual mean temperature and annual cumulative precipitation), and land use patterns (the fraction of agricultural land use and dam density). Using this data set, we assessed whether observed effects of the macro-scale factors are consistent across the two geographic regions by developing global and region-specific robust regression models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Materials and Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Materials and Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions. The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Materials**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual control of $\gamma$ diversity to emerge. Further, the consistent patterns across geographic regions with distinct fish fauna suggest that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

We used `r n_scenario` sets of parameter combinations (i.e., simulation scenarios) based on the results of extensive sensitivity analysis (**Supplementary Materials**). The parameter combinations comprises four landscape and eight ecological scenarios. Landscape scenarios distinguish spatial patterns of habitat heterogeneity, while ecological scenarios change ecological traits of simulated species. For landscape scenarios, we used four combinations of environmental variation at headwaters ($\sigma_h = 0.01, 1$) and the degree of local environmental noise ($\sigma_l = 0.01, 1$). The inequality of $\sigma_h$ and $\sigma_l$ controls the relationship between branching complexity and habitat heterogeneity (**Figures 1A and 1B**). For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

@@

We compiled fish data at two geographic regions: the Hokkaido island, Japan, and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish [@fukushimaModellingEffectsDams2007; @fukushimaHokkaidoFreshwaterFish2011], monitoring data at protected watersheds [@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors [@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire island. For the Midwest, we assembled fish community data collected by the Iowa Department of Natural Resources [@iowadepartmentofnaturalresourcesBiologicalAssessmentIowa2004], Illinois Environmental Protection Agency and Illinois Department of Natural Resources [@illinoisenvironmentalprotectionagencyIllinoisWaterMonitoring2014], Minnesota Pollution Control Agency [@minnesotapollutioncontrolagencyDevelopmentFishbasedIndex2014], and Wisconsin Department of Natural Resources [@wisconsindepartmentofnaturalresourcesGuidelinesAssessingFish2018]. We detailed procedures for initial data selection in **Supplementary Materials**.

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating asymptotic species richness (**Supplementary Materials**).

@@

For the supported model of $\gamma$ diversity, we estimated average predictive comparisons of watershed area and branching probability to compare their effect sizes. The average predictive comparison is an expected change of the response variable per a unit difference of the explanatory variable of interest [@gelmanAveragePredictiveComparisons2007]. Unlike other common metrics of effect size (e.g., standardized regression coefficients), this method provides intuitive yet rigorous estimates of effect size in models with non-linear variable transformations (see **Supplementary Materials** for further details). We performed statistical analysis using R version 4.0.2 [@rcoreteamLanguageEnvironmentStatistical2020].
